Hippocampal replay of experience at real world speeds

### 2021-05-13 Article.tex

The brain has the remarkable ability to store and retrieve representations of past events, allowing memories of the past to influence the future. These memory processes depend critically on the hippocampus, where rapid plasticity during an experience is thought to drive the initial encoding of events (Cite Eichenbaum and Cohen?). Subsequently, hippocampal "replay" of stored representations during both slow-wave sleep and periods of immobility during waking contributes to the longer term storage and updating of event memories in distributed hippocampal-neocortical circuits (REFs). Experiences can take place over many seconds, minutes, hours or days, and to be useful in guiding behavior, memories of these events must be stored in such a way that they can be retrieved quickly to inform upcoming choices. Consistent with this need for rapid storage, hippocampal replay events can express time compressed representations of past experience, activating regions outside the hippocampus and presumably driving plasticity in those regions to promote memory storage. \citep{CarrHippocampalreplayawake2011}.

@@

We find that spatially coherent events indeed have smaller HPD regions than events with fragmented dynamics. \FIG{4}A and \FIG{4}D show two example SWRs that are classified as having stationary and continuous dynamics, respectively. The most probable positions (as estimated by the model) at each time step in these SWRs is concentrated in a small portion of the track and correspondingly, the HPD region is small throughout the SWRs. In contrast, \FIG{4}C shows a SWR where the dynamics are fragmented and correspondingly, the most probable positions are more spatially diffuse and the HPD region is much larger. The HPD region size also provides insights into the moment-by-moment structure of each event, which can change over the time course of a SWR. An example of this change is shown in \FIG{4}B, where the HPD region is small for most of the SWR until the end, at which point the uncertainty of position becomes much higher, reflecting a transition from a spatially coherent to a spatially incoherent representation. Overall, when we examine the average HPD region size for each SWR, grouped by dynamic, we find a clear bimodal split between spatially coherent dynamics and spatially incoherent dynamics (\FIG{4}E). For the spatially coherent dynamics, the average HPD region for each SWR was much smaller than the spatially incoherent dynamics (median 24 cm, spatially coherent vs. median 238 cm, spatially incoherent, p=2.2e-16, one-sided Mann-Whitney-U). The HPD region for the unclassified SWRs was similarly large.

@@

% Figure 4
\begin{figure}
\includegraphics[width=0.80\linewidth]{figures/Figure4/Figure4_v6}
\caption{Validation of classification using the 95\% Highest Posterior Density. \textbf{(A-D)}  Examples of the 95 \% Highest Posterior Density. In each column: top panel: Probability of dynamic over time. Shading and labels indicate dynamic categories. Middle panel: Posterior distribution of estimated linear position over time. Magenta dashed line indicates animal's position. Bottom panel: HPD region size---the integral of the position bins with the top 95\% probable values. \textbf{(E)} Average 95\% HPD region size for each dynamic category. Median 24 cm for spatially coherent vs. median 238 cm for spatially incoherent, p=2.2e-16, one-sided Mann-Whitney-U}
\label{fig:4}

@@

We also verified that our results were very unlikely to be driven by the firing of interneurons, which are less likely to project to other brain areas and exhibit less spatial specificity than principal neurons in the hippocampus \citep{HangyaComplementaryspatialfiring2010, WilentDiscretePlaceFields2007}). First, we checked if requiring five or more tetrodes to have detected spikes during the SWR (up from requiring two tetrodes be active) would influence the proportions of dynamics, with the thought that a single tetrode with a high rate interneuron might result in stationary or stationary-continuous mixture dynamics during SWRs. We found that our results were robust to this increase in threshold (\FIGSUPP[5]{fig5sup2}A). Indeed, most of our events and most periods of dynamics within each event included spikes from upwards of 10 tetrodes (\FIGSUPP[5]{fig5sup3}B, \FIGSUPP[5]{fig5sup3}A). Next, we removed spikes that had spike widths of less than 0.3 ms to reduce the potential contribution of narrow-waveform inhibitory interneurons (Fox and Ranck https://pubmed.ncbi.nlm.nih.gov/1183529/). This had little effect: we still observed a high proportion of SWRs with stationary and stationary-continuous mixture dynamics (\FIGSUPP[5]{fig5sup2}B). We then repeated this analysis on clustered data for 9 of the 10 animals, excluding putative interneurons based on spike width and firing rate and requiring that at least three putative pyramidal cells be active during the SWR. Although there were many fewer events to examine, we still observed many stationary and stationary-continuous mixtures in SWRs (\FIGSUPP[5]{fig5sup2}D). Finally, if interneurons were not spatially specific and driving the slower dynamics, we would expect to see the posterior be less spatially concentrated. Instead, we found that these dynamics tended to have narrow posteriors (as characterized by 95\% HPD region size in \FIG{4}E). This provides further evidence that our results were not the solely the result of the firing of interneurons.

@@

A third possible concern is that burst firing from a single cell is driving the stationary dynamics. There are several lines of evidence against this. First, as mentioned above, most of our SWRs had spikes from multiple tetrodes, including those with stationary dynamics (\FIGSUPP[5]{fig5sup3}A), so it is unlikely that a single cell is driving these dynamics. Second, the stationary dynamics have a longer duration than one would expect from a burst alone (\FIG{5}B, 54 ms median duration, 38-74 ms quartiles vs. bursts which have a typical duration of 6-24 ms). Third, only ~50\% of the spikes had interspike intervals (ISIs) of less than 6 ms during the stationary, stationary-continuous, and continuous dynamics (\FIGSUPP[5]{fig5sup3}A), indicating that rapid bursts made up at most half of the observed spikes, and likely fewer given that these ISIs are computed for each tetrode. .

One last concern is that we have combined spiking information from several hippocampal subfields (CA1, CA2, and CA3), which may influence the types of dynamics we observed. However, CA1 and CA3 spiking is tightly coordinated within and across hemispheres during SWR replay \citep{CarrTransientSlowGamma2012} and multiple previous papers have combined recordings across subfields and obtained results similar to those that restricted their analyses to CA1 (CITE DIBA , KArlsson). Finally, we also verified that when we used only CA1 tetrodes to decode (\FIGSUPP[5]{fig5sup2}E), we get similar proportions of dynamics as when using all the tetrodes. Therefore, we feel our combination of spiking information from multiple subfields is appropriate.

@@

We hypothesize that this large set of events supports the ability to store and update  representations of individual places and snippets of experience. Previous findings demonstrated that SWRs with movement- or immobility-related activity engage specific patterns of spiking activity outside the hippocampus \citep{YuDistincthippocampalcorticalmemory2017, JadhavCoordinatedExcitationInhibition2016}. Thus, we would expect that the large set of events with slow dynamics would also engage specific patterns of activity outside the hippocampus, perhaps allowing animals to store or update the value of specific locations or short movement trajectories without having to generate full trajectories representing one of multiple ways to arrive at a location \citep{YuHippocampalcorticalinteraction2015}. Events that contain multiple dynamics (e.g. two continuous representations separated by a brief fragmented representation) might also help bind together experiences that occur in the same environment but at different places. Alternatively, it might be that downstream structures are attuned to these dynamics, and that each period of coherent hippocamapal dynamics drives its own set of responses elsewhere in the brain.

>>>>

The brain has the remarkable ability to store and retrieve representations of past events, allowing memories of the past to influence the future. These memory processes depend critically on the hippocampus, where rapid plasticity during an experience is thought to drive the initial encoding of events \citep{EichenbaumCanWeReconcile2014}. Subsequently, hippocampal "replay" of stored representations during both slow-wave sleep and periods of immobility during waking contributes to the longer term storage and updating of event memories in distributed hippocampal-neocortical circuits \citep{Joohippocampalsharpwave2018}(REFs). Experiences can take place over many seconds, minutes, hours or days, and to be useful in guiding behavior, memories of these events must be stored in such a way that they can be retrieved quickly to inform upcoming choices. Consistent with this need for rapid storage, hippocampal replay events can express time compressed representations of past experience, activating regions outside the hippocampus and presumably driving plasticity in those regions to promote memory storage. \citep{CarrHippocampalreplayawake2011}.

@@

We find that spatially coherent events indeed have smaller HPD regions than events with fragmented dynamics. \FIG{4}A and \FIG{4}B show two example SWRs that are classified as having continuous and stationary dynamics, respectively. The most probable positions (as estimated by the model) at each time step in these SWRs is concentrated in a small portion of the track and correspondingly, the HPD region is small throughout the SWRs. In contrast, \FIG{4}C shows a SWR where the dynamics are fragmented and correspondingly, the most probable positions are more spatially diffuse and the HPD region is much larger. The HPD region size also provides insights into the moment-by-moment structure of each event, which can change over the time course of a SWR. An example of this change is shown in \FIG{4}D, where the HPD region is small for most of the SWR until the end, at which point the uncertainty of position becomes much higher, reflecting a transition from a spatially coherent to a spatially incoherent representation. Overall, when we examine the average HPD region size for each SWR, grouped by dynamic, we find a clear bimodal split between spatially coherent dynamics and spatially incoherent dynamics (\FIG{4}E). For the spatially coherent dynamics, the average HPD region for each SWR was much smaller than the spatially incoherent dynamics (median 24 cm, spatially coherent vs. median 238 cm, spatially incoherent, p=2.2e-16, one-sided Mann-Whitney-U). The HPD region for the unclassified SWRs was similarly large.

@@

% Figure 4
\begin{figure}
\includegraphics[width=0.80\linewidth]{figures/Figure4/Figure4_v7}
\caption{Validation of classification using the 95\% Highest Posterior Density. \textbf{(A-D)}  Examples of the 95 \% Highest Posterior Density. In each column: top panel: Probability of dynamic over time. Shading and labels indicate dynamic categories. Middle panel: Posterior distribution of estimated linear position over time. Magenta dashed line indicates animal's position. Bottom panel: HPD region size---the integral of the position bins with the top 95\% probable values. \textbf{(E)} Average 95\% HPD region size for each dynamic category. Median 24 cm for spatially coherent vs. median 238 cm for spatially incoherent, p=2.2e-16, one-sided Mann-Whitney-U}
\label{fig:4}

@@

We also verified that our results were very unlikely to be driven by the firing of interneurons, which are less likely to project to other brain areas and exhibit less spatial specificity than principal neurons in the hippocampus \citep{HangyaComplementaryspatialfiring2010, WilentDiscretePlaceFields2007, JinnoNeuronalDiversityGABAergic2007}). First, we checked if requiring five or more tetrodes to have detected spikes during the SWR (up from requiring two tetrodes be active) would influence the proportions of dynamics, with the thought that a single tetrode with a high rate interneuron might result in stationary or stationary-continuous mixture dynamics during SWRs. We found that our results were robust to this increase in threshold (\FIGSUPP[5]{fig5sup2}A). Indeed, most of our events and most periods of dynamics within each event included spikes from upwards of 10 tetrodes (\FIGSUPP[5]{fig5sup3}B, \FIGSUPP[5]{fig5sup3}A). Next, we removed spikes that had spike widths of less than 0.3 ms to reduce the potential contribution of narrow-waveform inhibitory interneurons \citep{FoxLocalizationanatomicalidentification1975}. This had little effect: we still observed a high proportion of SWRs with stationary and stationary-continuous mixture dynamics (\FIGSUPP[5]{fig5sup2}B). We then repeated this analysis on clustered data for 9 of the 10 animals, excluding putative interneurons based on spike width and firing rate and requiring that at least three putative pyramidal cells be active during the SWR. Although there were many fewer events to examine, we still observed many stationary and stationary-continuous mixtures in SWRs (\FIGSUPP[5]{fig5sup2}D). Finally, if interneurons were not spatially specific and driving the slower dynamics, we would expect to see the posterior be less spatially concentrated. Instead, we found that these dynamics tended to have narrow posteriors (as characterized by 95\% HPD region size in \FIG{4}E). This provides further evidence that our results were not the solely the result of the firing of interneurons.

@@

A third possible concern is that burst firing from a single cell is driving the stationary dynamics. There are several lines of evidence against this. First, as mentioned above, most of our SWRs had spikes from multiple tetrodes, including those with stationary dynamics (\FIGSUPP[5]{fig5sup3}A), so it is unlikely that a single cell is driving these dynamics. Second, the stationary dynamics have a longer duration than one would expect from a burst alone (\FIG{5}B, 54 ms median duration, 38-74 ms quartiles vs. bursts which have a typical duration of 6-24 ms, \citep{RanckJrStudiessingleneurons1973, HarrisTemporalInteractionSingle2001, TroppSneiderDifferentialbehavioralstatedependence2006}). Third, only ~50\% of the spikes had interspike intervals (ISIs) of less than 6 ms during the stationary, stationary-continuous, and continuous dynamics (\FIGSUPP[5]{fig5sup3}A), indicating that rapid bursts made up at most half of the observed spikes, and likely fewer given that these ISIs are computed for each tetrode. .

One last concern is that we have combined spiking information from several hippocampal subfields (CA1, CA2, and CA3), which may influence the types of dynamics we observed. However, CA1 and CA3 spiking is tightly coordinated within and across hemispheres during SWR replay \citep{CarrTransientSlowGamma2012} and multiple previous papers have combined recordings across subfields and obtained results similar to those that restricted their analyses to CA1 \citep{DibaForwardreversehippocampal2007a, KarlssonAwakereplayremote2009}. Finally, we also verified that when we used only CA1 tetrodes to decode (\FIGSUPP[5]{fig5sup2}E), we get similar proportions of dynamics as when using all the tetrodes. Therefore, we feel our combination of spiking information from multiple subfields is appropriate.

@@

We hypothesize that this large set of events supports the ability to store and update  representations of individual places and snippets of experience. Previous findings demonstrated that SWRs with movement- or immobility-related activity engage specific patterns of spiking activity outside the hippocampus \citep{YuDistincthippocampalcorticalmemory2017, JadhavCoordinatedExcitationInhibition2016}. Thus, we would expect that the large set of events with slow dynamics would also engage specific patterns of activity outside the hippocampus, perhaps allowing animals to store or update the value of specific locations or short movement trajectories without having to generate full trajectories representing one of multiple ways to arrive at a location \citep{YuHippocampalcorticalinteraction2015}. Events that contain multiple dynamics (e.g. two continuous representations separated by a brief fragmented representation) might also help bind together experiences that occur in the same environment but at different places. Alternatively, it might be that downstream structures are attuned to these dynamics, and that each period of coherent hippocamapal dynamics drives its own set of responses elsewhere in the brain. We further note that recent results from a study of activity patterns during movement indicates that individual theta cycles can have different dynamics as well (CITE Pfeiffer paper), suggesting that the underlying sequences seen during replay and theta may overlap in their content and dynamics.
